#Public Health and Economic Consequences of Storms

##Data Processing

Download file, format date time
```{r, cache = TRUE}
url1 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("stormdata.csv.bz2")){download.file(url1, "stormdata.csv.bz2")}
stormdata <- read.csv("stormdata.csv.bz2")
url2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
if(!file.exists("info.pdf")){download.file(url2,"info.pdf")}
```

```{r format, cache=TRUE}
stormdata$BGN_DATE <- strptime(stormdata$BGN_DATE, "%m/%d/%Y %H:%M:%S")
year <- stormdata$BGN_DATE
year <- year$year
stormdata$year <- year + 1900
```

Processing
```{r processing, cache=FALSE}
library(dplyr)
dmgdata <- select(stormdata, year, STATE, EVTYPE, INJURIES, FATALITIES,
                   PROPDMG, CROPDMG)
```

##Results

```{r, cache=FALSE}
ymedian <- median(dmgdata$year)
dmgdata <- filter(dmgdata, year >= ymedian)
```
The full dataset includes entries from the years 1950 - 2011.  However, the 
median year is `r ymedian`.  This indicates that half of the data falls within
the last decade of six.  The analysis focuses on the data collected from the 
median year and later to limit the effect of increased record collection over
time.

```{r plot1, cache=FALSE}
library(ggplot2)

```
