#Public Health and Economic Consequences of Storms

##Abstract

This analysis uses data from the NOAA storm database to investigate which types
of weather events cause the greatest damage in terms of public health and 
economic losses.  Data form 2002 - 2011 was used as it was more complete than 
earlier records.  Injury and fatality data were used to evaluate public health
consequences, while crop and property damage figures were used for economic 
losses.  Storms were ranked according to the mean yearly total in each category
to determine the top 5 and these storms' yearly totals were plotted.  The 
highest injuries were recorded for tornados, excessive heat, hurricanes/ 
typhoons, thunderstorm wind and lightning.  The greatest number of fatalities 
were caused by tornados, excessive heat, flash floods, lightning and rip 
currents.  For both crop and property damage, the leading causes were flash 
floods, floods, hail, thunderstorm wind, and tornados.  As thunderstorms and 
tornados were consistently among the most damaging accross all parameters, 
they should be considered highly damaging to both public health and economic
factors.

##Data Processing

Download file, format date time
```{r, cache = TRUE}
url1 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("stormdata.csv.bz2")){download.file(url1, "stormdata.csv.bz2")}
stormdata <- read.csv("stormdata.csv.bz2")
url2 <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
if(!file.exists("info.pdf")){download.file(url2,"info.pdf")}
```

```{r format, cache=TRUE}
stormdata$BGN_DATE <- strptime(stormdata$BGN_DATE, "%m/%d/%Y %H:%M:%S")
year <- stormdata$BGN_DATE
year <- year$year
stormdata$year <- year + 1900
```

Processing
```{r processing, cache=FALSE}
library(dplyr)
dmgdata <- select(stormdata, year, STATE, EVTYPE, INJURIES, FATALITIES,
                   PROPDMG, CROPDMG)
tstmwind <- unique(grep("TSTM", dmgdata$EVTYPE, value = TRUE))
tstmwind <- tstmwind[tstmwind != 'NON-TSTM WIND' & tstmwind != 'NON TSTM WIND'
                     & tstmwind != "TORNADOES, TSTM WIND, HAIL"]
dmgdata$EVTYPE[dmgdata$EVTYPE %in% tstmwind] <- 'THUNDERSTORM WIND'
dmgdata$EVTYPE[dmgdata$EVTYPE == 'HEAT'] <- 'EXCESSIVE HEAT'
```
In order to determine the most damageing storm types in regard to public health
and property damage, data for injuries, fatalities, property and crop damage 
were selected.  Data was analyzed looking at the whole year to limit effects of 
seasonal change and determine overall impact so only the year of the event was 
kept for analysis.  This analysis focuses on the impact across the US, however
state designation was kept in the rearranged dataset as it may be useful in 
further analysis for more focused policy decisions.  Several factor levels
that distinguished between different strengths of thunderstorm wind were 
combined as the study aimed to rank general storm types.

```{r, cache=FALSE}
ymedian <- median(dmgdata$year)
dmgdata <- filter(dmgdata, year >= ymedian)
```
The full dataset includes entries from the years 1950 - 2011.  However, the 
median year is `r ymedian`.  This indicates that half of the data falls within
the last decade of six.  The analysis focuses on the data collected from the 
median year and later to limit the effect of increased record collection over
time.

##Results

###Ranking Storms by Damage Parameters

```{r totals, cache=FALSE}
library(reshape2)
dmgmelt <- melt(dmgdata, id.vars = c("year","STATE", "EVTYPE"))
dmgsum <- dcast(dmgmelt, year + EVTYPE ~ variable, sum)
```
First yearly sums were calculated for each storm type.  

```{r top damage, cache=FALSE}
meandmg <- summarise(group_by(dmgsum, EVTYPE), injury = mean(INJURIES), 
                   fatality = mean(FATALITIES), prop = mean(PROPDMG), 
                   crop = mean(CROPDMG))
topinj <- arrange(meandmg, desc(injury))[1:5,1]
topftl <- arrange(meandmg, desc(fatality))[1:5,1]
topprop <- arrange(meandmg, desc(prop))[1:5,1]
topcrop <- arrange(meandmg, desc(crop))[1:5,1]

```
In order to determine the most damaging storm types, the yearly sums were
averaged across the data for each parameter.  The storm types with the highest 
mean for each value were stored.

*Highest Injuries*
```{r, cache=FALSE}
c(sapply(topinj, as.character))
```

*Highest Fatalities*
```{r, cache=FALSE}
c(sapply(topftl, as.character))
```

*Highest Property Damage*
```{r, cache=FALSE}
c(sapply(topprop, as.character))
```

*Highest Crop Damage*
```{r, cache=FALSE}
c(sapply(topcrop, as.character))
```

###Yearly Public Health Consequences

```{r health data, cache=FALSE}
inj <- filter(dmgsum, EVTYPE %in% c(sapply(topinj, as.character)))
fat <- filter(dmgsum, EVTYPE %in% c(sapply(topftl, as.character)))
```
Each yearly sum for injuries and fatalities was plotted to give greater detail
as to how the storm rankings changed over the time period of analysis.

```{r health plot, cache=FALSE}
library(ggplot2)
library(gridExtra)
Injuries <- ggplot(inj, aes(x= year, y = INJURIES, col = EVTYPE)) +
        geom_line() +
        coord_cartesian(ylim = c(0,2000))+
        ylab("Total Injuries") +
        xlab("Year") +
        scale_color_manual(values = c("indianred", "darkgreen", "darkorange", 
                "turquoise", "magenta"), name = "Storm Type", labels = 
                c("Excessive Heat", "Hurricane/Typhoon*", "Lightning", 
                "Thunderstorm Wind*", "Tornado"))
       
Fatalities <- ggplot(fat, aes(x= year, y = FATALITIES, col = EVTYPE)) +
        geom_line() +
        coord_cartesian(ylim = c(0,250))+
        ylab("Total Fatalities") +
        xlab("Year") + 
        scale_color_manual(values = c("indianred", "darkgoldenrod4", "darkorange", 
                "blue3", "magenta"), name = "Storm Type", labels = 
                c("Excessive Heat", "Flash Flood*", "Lightning", "Rip Current*", 
                  "Tornado"))
        
        
grid.arrange(Injuries, Fatalities, ncol = 1, 
             top = "Yearly Total Public Health Consequences")
```

Storms marked wth an * are different between the injury and fatality rankings.
Beyond the storm types with the highest injury and fatality rates, the plot 
shows that excessive heat and tornados appear to have greater swings in the 
year to year impact on public health.  Tornado injury and fatality data for 
2011 was cut off from the plot as an outlier to show greater detail, however
the slope of the line was maintained.

###Yearly Economic Consequences


```{r economic data, cache=FALSE}
prop <- filter(dmgsum, EVTYPE %in% c(sapply(topprop, as.character)))
crop <- filter(dmgsum, EVTYPE %in% c(sapply(topcrop, as.character)))
```
Each yearly sum for crop and property damage was plotted to give greater detail
as to how the storm rankings changed over the time period of analysis.

```{r economic plot, cache=FALSE}
library(ggplot2)
library(gridExtra)
Property <- ggplot(prop, aes(x= year, y = PROPDMG, col = EVTYPE)) +
        geom_line() +
        ylab("Property Damage (US $)") +
        xlab("Year") +
        guides(color=guide_legend(title="Storm Type"))
       
Cropplot <- ggplot(crop, aes(x= year, y = CROPDMG, col = EVTYPE)) +
        geom_line() +
        ylab("Crop Damage (US $)") +
        xlab("Year") +
        guides(color=guide_legend(title="Storm Type"))
        
grid.arrange(Property, Cropplot, ncol = 1, 
             top = "Yearly Total Economic Consequences")
```

This plot shows crop and property damage over time.  Thunderstorm wind 
consistently caused the most property damage across the time period analyzed, 
while hail was generally the leader in crop damage.  Economic consequences
appear to follow more consistent trends than injury and fatality data.  Also,
crop damage peaked across all 5 causes in 2008, while property damage appears
to have generally increased across the time of analysis.